Opt-K

### 2013-06-13 paper/paper.tex
added some work on the analytic approximation

A significant challenge in the automated construction of Markov state
models is the choice of the number of microstates. Although classical Hamiltonian dynamics form a continuous-time Markov chain in $\mathbb{R}^{6N}$, the Markov property does not hold after the projecting the dynamics onto a basis of discrete indication functions -- the MSM microstates. In particular, when microstates contain within them free energy barriers of substantial magnitude, the validity of the Markov assumption begins to suffer considerably. While this source of modeling error can be addressed by increasing the number of microstates, the reduction in one error comes at the expense of the increase in another. This second source of error is statistical in origin. As the number of states in the model grows, so does the number of parameters required to specify the transition matrix. Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per
parameter, which makes the model liable for over-fitting.

@@

The proper likelihood is not of the trajectory within the discrete state
space; instead it is the likelihood of the trajectory within the
continuous phase space, on which the discrete states are merely an
indicator function basis.

@@

Up until this point, the emission probability distributions have not
been a model parameter for MSMs. Nevertheless, they are a critical
parameter for evaluating trajectory likelihoods. For example, consider two MSMs with the same transition probabilities. However, in one case, the emission distributions for each state are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the clusters. If the trajectory actually does go through the first models' regions of high likelihood, it would be clear that the first model more accurately fits the data.

However, the long timescale behavior of the MSM, the quantity of most
interest scientifically, is independent of the choice of the emission
distributions. It is only determined by the eigenspectrum of the
transition matrix elements. The emission distributions determine the
estimated eigenfunctions, but not the eigenvectors.

Therefore, the most appropriate emission distribution for discrete state
MSMs is that of the uniform distribution over the phase-space volume of
the state. That is, the likelihood of observing a conformation in phase
space given that the conformation is assigned to state $i$ is $0$
if the conformation is outside of the bounding volume of the state and
constant if the conformation is within the volume. This constant is set
so that the distribution integrates to $1$, and is thus the reciprocal
volume of the microstate.

@@

Briefly, we generated uniform random vectors within a bounding box whose size was slightly larger than the data. This random sample was then accepted if it was 'close enough' to a generator in the 100 state model, where close enough meant it was within 0.2 units from any generator in that model. This was used to approximate the convex hull of the data. Then the accepted points were assigned to a state. Then the volume of a state is proportional to the number of samples assigned to that state divided by the total number of samples. %We note, that these volumes are actually in units of the total acceptable volume. This affects the absolute value of the likelihood, however, since the total volume was constant in all cases, the relative likelihoods are the same.
\begin{figure}[h!]
\includegraphics[width=2in]{figs/muller_pot.png}
\caption{The M\"{u}ller potential consists of three metastable wells. We simulated Langevin dynamics on this potential as a validation of the new likelihood function.}
\label{fig:muller_pot}
\end{figure}

>>>>

A significant challenge in the automated construction of Markov state
models is the choice of the number of microstates. Although classical Hamiltonian dynamics form a continuous-time Markov chain in $\mathbb{R}^{6N}$, the Markov property does not hold after the projecting the dynamics onto a basis of discrete indication functions -- the MSM microstates. In particular, when microstates contain within them free energy barriers of substantial magnitude, the validity of the Markov assumption begins to suffer considerably. While this source of modeling error can be addressed by increasing the number of microstates, the reduction in one error comes at the expense of the increase in statistical error. As the number of states in the model grows, so does the number of parameters required to specify the transition matrix. Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per
parameter, which makes the model liable for over-fitting.

@@

The proper likelihood is not of the trajectory within the discrete state
space; instead it is the likelihood of the trajectory within the
continuous phase space on which the discrete states are merely an
indicator function basis.

@@

Up until this point, the emission probability distributions have not
been a model parameter for MSMs. Nevertheless, they are a critical
parameter for evaluating trajectory likelihoods. For example, consider two MSMs with the same transition probabilities. In one case, the emission distributions for each state are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the clusters. If the trajectory actually does go through the first models' regions of high likelihood, it would be clear that the first model more accurately fits the data.

However, the long timescale behavior of the MSM, the quantity of most
interest scientifically, is largely independent of the choice of the
emission distributions. %It is determined by the eigenspectrum of the
%transition matrix elements. The emission distributions determine the
%estimated eigenfunctions, but not the eigenvectors.

For this study, we consider the emission distribution of each state to be the 
uniform distribution over the phase-space volume of the state. That is, the 
likelihood of observing a conformation in phase space given that the 
conformation is assigned to state $i$ is $0$ if the conformation is outside of 
the bounding volume of the state and constant if the conformation is within 
the volume. This constant is set so that the distribution integrates to $1$, 
and is thus the reciprocal volume of the microstate.

@@

Briefly, we generated uniform random vectors within a bounding box whose size was slightly larger than the data. This random sample was then accepted if it was ``close enough'' to a generator in the 100 state model, where close enough meant it was within 0.2 units from any generator in that model. This was used to approximate the convex hull of the data. Then the accepted points were assigned to a state. Then the volume of a state is proportional to the number of samples assigned to that state divided by the total number of samples. %We note, that these volumes are actually in units of the total acceptable volume. This affects the absolute value of the likelihood, however, since the total volume was constant in all cases, the relative likelihoods are the same.
\begin{figure}[h!]
\includegraphics[width=2.5in]{figs/muller_pot.png}
\caption{The M\"{u}ller potential consists of three metastable wells. We simulated Langevin dynamics on this potential as a validation of the new likelihood function.}
\label{fig:muller_pot}
\end{figure}
